Comparing ChatGPT-4 to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6942 Deaths in Sierra Leone from 2019-2022

### 2024-01-24 manuscript/manuscript.tex
Add minimal interva and insilicova code

\textbf{Methods:} 6942 records collected from VA in Sierra Leone from 2019 to 2022 were used to compare four computer models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, to physicians for assigning CODs at the population and individual level. Open narratives were used for Chat-3.5/ChatGPT-4 input, while structured questionnaires were used for InterVA-5/InSilicoVA input. All COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for the adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at the population and individual level respectively. Comparisons in CSMF and PCCC to physicians among models were evaluated for all records and by COD, age group, and age ranges.

@@

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \cite{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. On November 30, 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \cite{wuBriefOverviewChatGPT2023}. On March 14, 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \cite{openaiGPT4TechnicalReport2023}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conduct a case study with Sierra Leone deaths from VA in 2019 to 2022 to compare four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

The performances of four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, for determining CODs were evaluated using 6942 physician agreed records in 2019 to 2022 from the Health Sierra Leone (HEAL-SL) study \cite{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022}. Physician agreed records are dual-coded records where two physicians had similar COD assignments. COD outputs from the four models were compared to CODs assigned by physicians.

\subsection{Healthy Sierra Leone (HEAL-SL) Study Dataset}\label{methods-healsl}

The HEAL-SL study aims to \cite{carshon-marshChildMaternalAdult2022}.

@@

The initial 11,920 records were collected from dual-coded Electronic Verbal Autopsy (EVA), where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \cite{worldhealthorganizationICD10InternationalStatistical2011}. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). To determine if two codes were in physician agreement per record, codes were compared for similarity using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which groups a range of similar ICD-10 codes together indicating codes in agreement \cite{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 1). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and initial codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes were not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes before and after reconciliation, and assigns a final code based on their evaluation. The 11,820 physician coded records were further filtered for records where both physicians agreed on the assigned codes (records that were not reconciled or adjudicated), resulting in 6942 physician agreed records. Since computer models were compared to physicians in this study, there was more certainty that COD assignments agreed by both physicians were representative of physician assignment than when they disagreed \cite{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}.

@@

Four computer models were used to predict COD for each of the 6942 physician agreed records: ChatGPT-3.5, ChatGPT-4, InterVA-5 , and InSilicoVA. Each model required pre-processing of the 6942 physician agreed records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from ChatGPT-3.5 and ChatGPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation.

@@

ChatGPT-3.5 \cite{brownLanguageModelsAre2020} and ChatGPT-4 \cite{openaiGPT4TechnicalReport2023} are LLMs that utilize deep neural networks with transformer architectures \cite{vaswaniAttentionAllYou2017} and reinforcement learning from human feedback \cite{ouyangTrainingLanguageModels2022,christianoDeepReinforcementLearning2017,stiennonLearningSummarizeHuman2020,wirthSurveyPreferencebasedReinforcement2017} to follow instructions from prompts and provide human-level responses, with known differences in ChatGPT-4 possessing multimodal capabilites (e.g. image/voice input/output), more recent training data, and improved responses compared to ChatGPT-3 \cite{wuBriefOverviewChatGPT2023}. ChatGPT-3.5 and ChatGPT-4 were instructed with text prompts to assign CODs based on the open narrative. InterVA-5 and InSilicoVA are widely used and studied \cite{murrayUsingVerbalAutopsy2014,otiVerbalAutopsyInterpretation2010,benaraEvaluationMethodsAssigning2023,chandramohanEstimatingCausesDeath2021,tungaVerbalAutopsyModels2021,leitaoComparisonPhysiciancertifiedVerbal2014,jhaAutomatedPhysicianAssignment2019} standard statistical models for COD assignment in VAs under the openVA framework \cite{liOpenVAToolkitVerbal2023}. InterVA-5 applies Bayesian probabilistic modelling \cite{bayesEssaySolvingProblem1958} using a set of standardized symptoms from reports and related conditional probabilities from medical experts to assign CODs based on the highest probability \cite{byassIntegratedApproachProcessing2019,byassStrengtheningStandardisedInterpretation2012}. InSilicoVA improves upon InterVA (e.g. comparable probabilities across individuals, measures of uncertainty, and inclusion of additional data sources) with a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) simulations \cite{brooksMarkovChainMonte1998,chibMarkovChainMonte2001,hanMarkovChainMonte2001} to incorporate multiple sources of uncertainty for assigning CODs based on the highest probability \cite{mccormickProbabilisticCauseofDeathAssignment2016b}. For assigning CODs, ChatGPT-3.5 and ChatGPT-4 require prompts containing conversation-like textual instructions as input, while InterVA-5 and InSilicoVA require structured symptom and associated sociodemographic data as input.

@@

Of the 6942 physician agreed records, ChatGPT-3.5 and ChatGPT-4 were able to assign CODs for 6932 and 6912 records respectively, while InterVA-5 and InSilicoVA both assigned CODs for 6833 records. All 6833 records with WHO VA 2016 v1.5 codes \cite{worldhealthorganizationVerbalAutopsyStandards2016} from InterVA-5 and InSilicoVA output were converted into ICD-10 codes. After all model outputs contained ICD-10 codes, they were further converted to Centre for Global Health Research 10 (CGHR-10) codes that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 years or older), child (28 days to 11 years), and neonatal (under 28 days) age groups (see Additional file 2). The 6932 and 6912 records with ICD-10 codes from ChatGPT-3.5 and ChatGPT-4 output were converted into CGHR-10 codes, resulting in 6894 and 6857 records, where 38 and 55 records did not have matching CGHR-10 codes respectively. The 6833 records with ICD-10 codes from InterVA-5 and InSilicoVA output were also converted into CGHR-10 codes, resulting in and 6805 and 6729 records, where 28 and 104 records did not have matching CGHR-10 codes respectively.

@@

The performance of four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, were evaluated with metrics on the population and individual level by comparing their CGHR-10 COD outputs to 6942 physician agreed records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level, while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level \cite{murrayRobustMetricsAssessing2011}. Records that were assigned a COD by physicians, but not by a model were considered to be an incorrect COD assignment by the model. CSMF accuracy and PCCC were calculated for each model by three age groups, and further into age ranges and cause of death for each age group.

>>>>

\textbf{Methods:} 6939 records collected from VA in Sierra Leone from 2019 to 2022 were used to compare four computer models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, to physicians for assigning CODs at the population and individual level. Open narratives were used for Chat-3.5/ChatGPT-4 input, while structured questionnaires were used for InterVA-5/InSilicoVA input. All COD assignments were grouped into general COD categories consisting of 19, 10, and 7 categories for the adult, child, and neonatal age groups. Cause Specific Mortality Fraction (CSMF) accuracy and Partial Corrected Concordance (PCCC) were used to compare models to physicians at the population and individual level respectively. Comparisons in CSMF and PCCC to physicians among models were evaluated for all records and by COD, age group, and age ranges.

@@

Recently, Large Language Models (LLM), leveraging massive datasets and deep learning approaches, have made advances in performing a variety of Natural Language Processing (NLP) tasks using free-text, such as question answering, code generation, and even medical diagnosis \cite{changSurveyEvaluationLarge2023,lundChattingChatGPTHow2023,svyatkovskiyIntelliCodeComposeCode2020,hauptAIGeneratedMedicalAdvice2023}. On November 30, 2022, a widely-available LLM called ChatGPT was released by OpenAI with capabilities of answering natural language text inquiries using training data up to September 2021. ChatGPT-3 was based on several Generative Pre-trained Transformer (GPT) models between 2018 to 2020, namely GPT-1 to GPT-3, which had notable differences in training data sizes of 5 gigabytes to 45 terabytes from web sources that resulted in 117 million to 175 billion parameter models \cite{wuBriefOverviewChatGPT2023}. On March 14, 2023, ChatGPT-4 was released with human-level performance on various professional and academic exams and benchmarks that outperformed ChatGPT-3 \cite{openaiGPT4TechnicalReport2023a}. Given the limited usage of free-text open narratives in computer models for determining CODs, and recent advances in LLMs that leverage natural language text prompts, we conduct a case study with Sierra Leone deaths from VA in 2019 to 2022 to compare four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, to physicians for determining CODs.

@@

This study uses 6939 physician agreed records of 11,920 records collected in 2019 to 2022 from the Healthy Sierra Leone (HEAL-SL) as described in Section \ref{methods-phyagree}. Section \ref{methods-models} describes the four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, for COD assignment, and their inputs and outputs, while section \ref{methods-eval} details the performance evaluation of the four models relative to physicians overall, by ages, and by COD, using population and individual level metrics.

@@

Initially, 11,920 records from the HEAL-SL study \cite{njalauniversityHealthySierraLeone2023,carshon-marshChildMaternalAdult2022} were collected from dual-coded EVA, where each record was randomly coded by two different physicians that assigned CODs as International Classification of Diseases Revision 10 (ICD-10) codes \cite{worldhealthorganizationICD10InternationalStatistical2011}. Physicians were able to assign CODs for 11,820 of the 11,920 records, where 100 of these records could not be assigned a COD due to missing or inadequate information (e.g. low quality narrative, data loss). To determine if two codes were in physician agreement per record, codes were compared for similarity using Central Medical Evaluation Agreement 10 (CMEA-10) codes, which groups a range of similar ICD-10 codes together indicating codes in agreement \cite{aleksandrowiczPerformanceCriteriaVerbal2014} (see Additional File 1). When codes were not in agreement, a record enters the reconciliation phase, where the two physicians were provided reasoning and initial codes from each other to: (1) keep their initial code (2) assign the other physician's code or (3) assign a new code. If codes were not in agreement after the reconciliation phase, a record enters the adjudication phase, where a third senior physician evaluates both physicians' reasoning and codes before and after reconciliation, and assigns a final code based on their evaluation. The 11,820 physician coded records were further filtered for records where both physicians agreed on the assigned codes (records that were not reconciled or adjudicated), resulting in 6942 physician agreed records. Since computer models were compared to physicians in this study, there was more certainty that COD assignments agreed by both physicians were representative of physician assignment than when they disagreed \cite{barnettComparativeAccuracyDiagnosis2019,morrisFactorsAssociatedPhysician2010,hsiaoFactorsAssociatedPhysician2012}.

@@

Four computer models were used to predict COD for each of the 6942 physician agreed records: ChatGPT-3.5, ChatGPT-4, InterVA-5 , and InSilicoVA. Each model required pre-processing of the 6942 physician agreed records into input data, and standardization of output COD codes from models for performance evaluation as not all models produced comparable codes across outputs. Although each model can assign multiple CODs per record, only the first generated COD response from ChatGPT-3.5 and ChatGPT-4, and the most probable COD from InterVA-5 and InSilicoVA were used for evaluation. After conversion of all model and physician outputs to generalized Centre for Global Health Research 10 (CGHR-10) codes (see Additional File 2), 6939 records were comparable to evaluate performances of models for COD assignment relative to physicians.

@@

ChatGPT-3.5 \cite{brownLanguageModelsAre2020} and ChatGPT-4 \cite{openaiGPT4TechnicalReport2023a} are LLMs that utilize deep neural networks with transformer architectures \cite{vaswaniAttentionAllYou2017} and reinforcement learning from human feedback \cite{ouyangTrainingLanguageModels2022,christianoDeepReinforcementLearning2017,stiennonLearningSummarizeHuman2020,wirthSurveyPreferencebasedReinforcement2017} to follow instructions from prompts and provide human-level responses, with known differences in ChatGPT-4 possessing multimodal capabilites (e.g. image/voice input/output), more recent training data, and improved responses compared to ChatGPT-3 \cite{wuBriefOverviewChatGPT2023}. ChatGPT-3.5 and ChatGPT-4 were instructed with text prompts to assign CODs based on the open narrative. InterVA-5 and InSilicoVA are widely used and studied \cite{murrayUsingVerbalAutopsy2014,otiVerbalAutopsyInterpretation2010,benaraEvaluationMethodsAssigning2023,chandramohanEstimatingCausesDeath2021,tungaVerbalAutopsyModels2021,leitaoComparisonPhysiciancertifiedVerbal2014,jhaAutomatedPhysicianAssignment2019} standard statistical models for COD assignment in VAs under the openVA framework \cite{liOpenVAToolkitVerbal2023}. InterVA-5 applies Bayesian probabilistic modelling \cite{bayesEssaySolvingProblem1958} using a set of standardized symptoms from reports and related conditional probabilities from medical experts to assign CODs based on the highest probability \cite{byassIntegratedApproachProcessing2019,byassStrengtheningStandardisedInterpretation2012}. InSilicoVA improves upon InterVA (e.g. comparable probabilities across individuals, measures of uncertainty, and inclusion of additional data sources) with a hierarchical Bayesian framework and Markov Chain Monte Carlo (MCMC) simulations \cite{brooksMarkovChainMonte1998,chibMarkovChainMonte2001,hanMarkovChainMonte2001} to incorporate multiple sources of uncertainty for assigning CODs based on the highest probability \cite{mccormickProbabilisticCauseofDeathAssignment2016b}. For assigning CODs, ChatGPT-3.5 and ChatGPT-4 require prompts containing conversation-like textual instructions as input, while InterVA-5 and InSilicoVA require structured symptom and associated sociodemographic data as input.

@@

Of the 6942 physician agreed records, ChatGPT-3.5 and ChatGPT-4 were able to assign CODs for 6932 and 6912 records respectively, while InterVA-5 and InSilicoVA both assigned CODs for 6833 records. All 6833 records with WHO VA 2016 v1.5 codes \cite{worldhealthorganizationVerbalAutopsyStandards2016} from InterVA-5 and InSilicoVA output were converted into ICD-10 codes. After all model outputs contained ICD-10 codes, they were further converted to CGHR-10 codes that generalized ICD-10 codes into 19, 10, and 7 categories for the adult (12 years or older), child (28 days to 11 years), and neonatal (under 28 days) age groups (see Additional file 2). The 6932 and 6912 records with ICD-10 codes from ChatGPT-3.5 and ChatGPT-4 output were converted into CGHR-10 codes, resulting in 6894 and 6857 records, where 38 and 55 records did not have matching CGHR-10 codes respectively. The 6833 records with ICD-10 codes from InterVA-5 and InSilicoVA output were also converted into CGHR-10 codes, resulting in and 6805 and 6729 records, where 28 and 104 records did not have matching CGHR-10 codes respectively. The 6942 physician agreed records with COD assignments in ICD-10 were also converted to CGHR-10 codes, which resulted in 6939 physician agreed records for comparison across physician and model assignments (three records in ICD-10 were not captured by the CGHR-10 ranges).

@@

The performance of four models, ChatGPT-3.5, ChatGPT-4, InterVA-5, and InSilicoVA, were evaluated with metrics on the population and individual level by comparing their CGHR-10 COD outputs to 6939 physician agreed records. Cause Specific Mortality Fraction (CSMF) accuracy was used to evaluate models on the population level, while Partial Chance Corrected Concordance (PCCC) was used to evaluate models on the individual level \cite{murrayRobustMetricsAssessing2011}. Records that were assigned a COD by physicians, but not by a model were considered to be an incorrect COD assignment by the model. CSMF accuracy and PCCC were calculated for each model by three age groups, and further into age ranges and cause of death for each age group.
